---
title: "CherryBlssm_Pred"
author: "Marie Han"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
```

Loading training data
```{r}
cherry <- read.csv("data/washingtondc.csv") |> 
  bind_rows(read.csv("data/liestal.csv")) |> 
  bind_rows(read.csv("data/kyoto.csv")) |> 
  bind_rows(read.csv("data/vancouver.csv")) |> 
  bind_rows(read.csv("data/nyc.csv"))
```



Function to obtain weather data:
```{r}
get_ghcn_station= function(station_id, destdir = tempdir()) {
  base_url= "https://www.ncei.noaa.gov/pub/data/ghcn/daily/by_station/"
    file_name= paste0(station_id, ".csv.gz")
    url= paste0(base_url, file_name)
    destfile= file.path(destdir, file_name)
    if (!file.exists(destfile)) {
      download.file(url, destfile, mode = "wb")
    }
    data=read.csv(destfile, header = FALSE)
    data=data[, -c(5:8)]
    data$Date=as.Date(as.character(data$V2), "%Y%m%d")
    data$Year=as.integer(format(data$Date, "%Y"))
    
    #Pivot to get vars as columns
    data_wide=data %>%
      select(Date, Year, V3, V4) %>%
      pivot_wider(
        names_from = V3,
        values_from = V4)
    
    # Calculate TAVG if TMAX and TMIN exist
    if (all(c("TMAX", "TMIN") %in% colnames(data_wide))) {
      data_wide= data_wide %>%
        mutate(TAVG = (TMAX + TMIN)/2)
    }
    
    # Keep only the desired elements
    elements_to_keep= c("TMIN", "TMAX", "TAVG", "PRCP")
    data_wide= data_wide %>%
      select(Date, Year, any_of(elements_to_keep))
    
    # Remove rows without both TMIN and TMAX
    data_wide= data_wide %>%
      filter(!is.na(TMIN) & !is.na(TMAX))
    
    return(data_wide)
}

```


Weather by location, putting it into one "daily" dataframe.
```{r}


kyototemp=get_ghcn_station("JA000047759")
washdc=get_ghcn_station("USW00013743")
vancouvertemp=get_ghcn_station("CA001108395")
newyorkcity=get_ghcn_station("USW00014732")
liestaltemp=get_ghcn_station("SZ000001940")

#Adding a location variable for identification
kyototemp$location <- "kyoto"
liestaltemp$location <- "liestal"
newyorkcity$location <- "newyorkcity"
vancouvertemp$location <- "vancouver"
washdc$location <- "washingtondc"

#Now putting them all into one file
Cherry_day <- bind_rows(kyototemp,liestaltemp, newyorkcity, vancouvertemp, washdc) %>%
  rename(year = Year)%>% #to match with the original cherry year label 
  mutate(Date = as.Date(Date),
         year = as.numeric(format(Date, "%Y")),
         month = as.numeric(format(Date, "%m")),
         bloom_year = ifelse(month >= 10, year+1, year)
         )
```
Adding precipitation counter, 
I wanted to capture the rain accumulation over a week, as an inch per week is considered ideal for cherry tree growth and health. This is also whhere I had to start collapsing each individual day's min/max/average temperature, in order to start looking at things more broadly. Because this is only looking at data starting in October until February,  I was not expecting a lot of really hot weather to begin with, but I wasn't counting out the possibility. 
```{r}
#To filter things weekly ###
Cherry_week <- Cherry_day %>%
  filter(month >= 10 | month <= 2) %>% 
  mutate(week = isoweek(Date)
  ) %>%
  group_by(location, bloom_year, week) %>%
  summarise(
    weekly_PRCP = sum(PRCP),
    mean_TAVG   = mean(TAVG),
    min_TAVG    = min(TAVG),
    max_TAVG    = max(TAVG),
    .groups = "drop"
  ) %>%
  mutate(
    PRCP_indicator    = ifelse(weekly_PRCP > 254, 1, 0),  # > 254 tenths of mm, as is meausured by data set = 1 inch
    TooCold_indicator = ifelse(min_TAVG < -29, 1, 0), #more important if it's been too cold for an entire week than just a singular day 
    TooHot_indicator  = ifelse(max_TAVG > 29, 1, 0)
  )



```

Now to make our data usable in the model and prediction, collapsing and summarizing it into a year in order to match the original "cherry" format. 
```{r}
Cherry_Year <- Cherry_week %>%
  group_by(location,bloom_year) %>%
  summarize(
    precipitation = sum(PRCP_indicator),
    avg_TMAX = mean(max_TAVG),
    avg_TMIN = mean(min_TAVG),
    AVG = mean(mean_TAVG),
    TooCold = sum(TooCold_indicator),
    TooHot = sum(TooHot_indicator),
    .groups = "drop"
) %>%
  #select(-year) %>%
  rename(year = bloom_year ) %>%
  mutate(precipitation = replace_na(precipitation, 0))

```



Model

The too hot indicator was insignificant due to it being fall and winter time during our windowed time frame. It's very unlikely that any of these locations would have a week or more long heatwave, but there is always the possibility. 

```{r}
model1_TEST <- cherry |>
  inner_join(Cherry_Year,
             by = c("location", "year")) |>
  lm(formula = bloom_doy ~ year * location +  AVG + TooCold+TooHot +precipitation )
  
  
  

summary(model1_TEST)
```



```{r}
###TEST PREDICTIONS

cherry_gridTEST <- expand_grid(location = unique(cherry$location),
                           year = 1990:2026) |> 
  inner_join(Cherry_Year,
             by = c("location", "year"))

predictions2 <- cherry_gridTEST |> 
  mutate(pred_bloom = predict(model1_TEST, newdata = cherry_gridTEST))

predictions2 |> 
  left_join(cherry,
             by = c("location", "year")) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = pred_bloom)) +
  facet_grid(rows = vars(location))

```


## Submission

```{r}

predictions2 |> 
  filter(year == 2026) |>
  mutate(predicted_date= strptime(paste(year,pred_bloom), "%Y %j") |>
           as_date())
```

__Predicted bloom dates__

DC: 04/03/2026

Liestal: 04/02/2026

Kyoto: 04/01/2026

NYC: 04/08/2026
